---
title: "R for Planning. Week One"
author: "Henry Renski"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week One

### Exploring R for Data Analysis and Visualization

Our objective for today is to simply get a feel for R and the R Studio interface. Today we will run through some of the basic features in R in what is a very common type of project workflow. In today's exercise we will do the following:

1.  Explore the R Studio interface

2.  Install and load the packages

3.  Download data from American Community Survey through the Census Bureau API server

4.  Format the data and calculate some summary statistics

5.  Make a map showing the spatial distribution of some variables

6.  Make and format some descriptive charts

I'll be proving the code for you to run. Its OK if you don't understand what we are doing. Over the course of the semester, we will be exploring these (and other) functions in much greater detail.

Note: I have written these instructions is a package called R Markdown, which makes nice documents that allow you to interweave narrative, code and charts. This is referred to as literate coding or literate programming.

### Explore the R Studio Interface

Let's look at the four windows in R.

The gray chunk below is a code block. You can issue simple commands into the console, and its where you will find messages and some printed output. I want you to copy and paste the code below into the console

```{r}
1+1


```

Open a new "R script" file.

### Install and load packages

While everything can be done in R's base code, most people use a variety of "packages" for analysis. Packages are blocks of code written by other users. You have to install the packages (and then load) the you want to use each time you run code. There are thousands upon thousands of packages in R, and they all have their on syntax and grammar. However, there are some packages that you wind up using all the time, and those once you master those few you will be fine.

Let's load the packages you will need for today.

Open a new "R script" file. Then copy and paste the following code into the file. Then highlight and run the code.

```{r install, eval=FALSE, message=FALSE, warning=TRUE}

#code below to install packages
install.packages("tidyverse")  #suite including many of the best data munipilation and visualization functions

install.packages("tidycensus")  #pulls ACS data off the census bureau API, works well with dplyr and ggplot but doesn't cover all census data products

install.packages("tigris")      #pulls GIS files directly from census TIGER server, integrated with tidycensus

```

Packages typically only need to be installed one time (per computer). However, your desired packages will need to be "loaded" each time you start a new R session. The next code chunk loads the required packages.

```{r libraries, echo=TRUE, message=FALSE, warning=FALSE}

library(tidyverse)
library(tidycensus)
library(tigris)

```

We can see if the package was loaded in the lower right window and choosing the "Packages" tab.

### Accessing Data through the Census API

Rather than download data onto your computer, why not work with it directly from the Census Bureau's servers? That's the point behind working with an API. We will use the **tidycensus** package to access the American Community Survey API and the associated **tigris** package to access the spatial files (for mapping).

To get started working with **tidycensus**, users must first get a Census API key. The API key is like a password that you need to access census data. A key can be obtained from [\<http://api.census.gov/data/key_signup.html\>](http://api.census.gov/data/key_signup.html){.uri}. You only need to sign up for it once, as long as you store the code in a place you remember for future use.

```{r APIkey, eval=FALSE, message=FALSE, warning=FALSE}

census_api_key("YOUR API KEY GOES HERE")

# census_api_key("9eac35e5489829fedcaaf690f14d2ec112dd69ab")
```

Let's get data on household income data for Massachusetts Census Tracts using the TidyCensus Package to access the Census API. You don't even need to save it to your computer. It runs straight off the Census Bureau API servers.

Note: TidyCensus is just one of many packages. It makes getting Census data a snap, but it only covers a small number of census data products. The censusapi package is more comprehensive, but more difficult to use.

```{r API, eval=FALSE, message=FALSE, warning=FALSE}
MA.tract.inc <- get_acs(geography = "tract", variables = "B19013_001",
                state = "MA", geometry = TRUE)

```

The geometry=TRUE option downloads spatial files along with the income data.

Let's look at the data set from the environment window in R studio.

### Calculating Summary Statistics using dplyr (part of tidyverse)

Let's calculate the mean and median of tract-level income. We will then store these variables as objects (called mean and median) for future use

```{r summary, eval=FALSE, message=FALSE, warning=FALSE}
mean <- as.numeric(as.data.frame(MA.tract.inc) %>% 
  summarise(mean_inc = mean(estimate, na.rm=TRUE)))

mean <- round(mean,digits = 0)

median <- as.numeric(as.data.frame(MA.tract.inc) %>% 
  summarise(mean_inc = median(estimate, na.rm=TRUE)))

```

Note that the mean and median have now been added as "Values" in the environment. 

### Making Maps using ggplot2 with tigris

Let's map the data! This works because the tidycensus package allows us to grab the corresponding spatial data from Tiger (via the tigris package). Most times we have to download the spatial data separately and join it ourselves (just like in GIS).

```{r map, eval=FALSE, message=FALSE, warning=FALSE}

ggplot(MA.tract.inc, aes(fill = estimate, color = estimate)) +
   geom_sf() +
   coord_sf(crs = 26987) +
   scale_fill_gradient(low="#E2F8FB",high="#0A4249") +
   scale_color_gradient(low="#E2F8FB",high="#0A4249") +
   labs(title="Median Household Income by Census Tract")

```

If we have time later we can play with some different color schemes using an online color wheel.

### Making Charts and Graphs using ggplot2

For our final exploration, let's create a histogram that shows distribution of tract-level incomes.  We will also annotate the chart with the overall mean and median values we calculated earlier.

```{r chart, eval=FALSE, message=FALSE, warning=FALSE}

options(scipen=999)  ##removes scientific notion

ggplot(MA.tract.inc, aes(x = estimate)) + geom_histogram(bins=16, fill="#1AB3C7", color="#0C5059") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_discrete(name ="Median Household Income", 
                    limits=c(0,50000,100000,150000,200000,250000)) +
    labs(title="Distribution of Median Household Incomes by Census Tract",
         y="Number of Tracts") +
    annotate("text", x=200000, y=c(300,280), 
             label=c(paste("Mean Income = ", mean),paste("Median Income = ", median)))
                                                  
```

That's all for our quick Intro to R. 

Don't worry if you feel lost. In the weeks ahead we will go into more detail and depth on how to use R for data analysis.
